---
title: "formula-syntax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{formula-syntax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(criticalESvalue)
```

In this vignette we illustrate a problem when using the `critical()` function with some models. In particular, all models of class `htest` (e.g., `t.test`, `cor.test`, etc.) do not store the dataset within the output list. On the other side, the `lm` function for example return a list of class `lm` that contains the dataset (`$model`).

```{r}
fit_ttest <- t.test(mpg ~ am, data = mtcars)
fit_lm <- lm(Sepal.Length ~ Petal.Width, data = iris)

class(fit_ttest)
class(fit_lm)

str(fit_ttest, max.level = 1)
head(fit_lm$model)
```

The `criticalESvalue`, in particular the `critical.*` functions use the `insigth::get_data()` function to retrieve the source dataset from the environment or the output list.

This works perfectly when the object contains the dataset as in `lm`:

```{r}
head(insight::get_data(fit_lm))
```

However, in the `htest` class the function sometimes fails. In particular, when using e.g., `t.test` with the formula syntax `y ~ x` and the `data = ` argument, `insight::get_data()` is not able to retrieve the dataset and return `NULL`. We included an error message suggesting to change the function call.

```{r, error=TRUE}
insight::get_data(fit_ttest)
critical(fit_ttest)
```

To fix this problem, you can simply change the function call. There are several options:

```{r}
# formula without data
fit_ttest_1 <- t.test(mtcars$mpg ~ mtcars$am)

# x and y without data
fit_ttest_2 <- t.test(mtcars$mpg[mtcars$am == 0], mtcars$mpg[mtcars$am == 1]) 

# creating variables in the global environment, not recommended

x <- mtcars$mpg[mtcars$am == 0]
y <- mtcars$mpg[mtcars$am == 1]
fit_ttest_3 <- t.test(x, y)
```

With these options, the `critical()` function will not fail:

```{r}
critical(fit_ttest_1)
critical(fit_ttest_2)
critical(fit_ttest_3)
```

If we found a stable and reliable workaround to handle `htest` objects, this behaviour could change in future versions of the package.